---
title: "TWAS-GSEA-fast Tier-A calibration & power study"
author: "sim/ harness"
date: "2026-09-11"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: kable
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE)
suppressMessages({library(data.table); library(knitr)})
```

# Introduction

`TWAS-GSEA-fast.R` (branch `optimisation`) performs competitive gene-set / gene-property
enrichment on TWAS results. The statistical model is a two-variance-component LMM

$$ y = X\beta + g + e, \qquad g \sim \mathcal{N}(0, \sigma_u^2 K), \qquad e \sim \mathcal{N}(0, \sigma_e^2 I) $$

where $K$ is the block-diagonal predicted-expression correlation matrix produced by
`build_cor_matrix.R`. `K` is currently constructed as `abs(WGCNA::cor(...))` (line 40 of
`R/build_cor_matrix_helper.R`) — signed correlations are collapsed to magnitudes. The
question this study asks is: **is the abs-K default correctly calibrated, and does replacing
it with a signed-K variant improve power on directional properties without breaking Type-I
error?**

To answer that we need a simulator that

1. draws TWAS Z under a realistic block-diagonal covariance,
2. supports null (mu = 0) and signal (mu = $\beta t$) injection modes,
3. runs the tool (or a bit-identical in-session port of it) fast enough to accumulate
   thousands of replicates.

The harness lives under `sim/`. Everything downstream reproduces from this document once
the four phase drivers below have run.

## Environment

All R scripts are called through a wrapper `sim/run_r.sh` that invokes the `twas_gsea`
conda env's Rscript with `env -i`, because the container's `~/.Renviron` hardcodes
`R_LIBS_USER` to an R-4.2 user library that ABI-clashes with the env's R 4.0.

```{bash env}
# One-off: create the conda env from the checked-in spec.
micromamba env create -f env.yaml   # r-base=4.0.2, r-wgcna, r-vgam, r-qusage, ...

# Wrapper (`sim/run_r.sh`) sets HOME to a scratch dir with no .Renviron and forces
# R_LIBS_USER to a nonexistent path so only the env library is loaded.
```

Every code chunk below that starts an R script goes through `./sim/run_r.sh`.

# Phase 1: build the correlation matrices

Two matrices, one abs-K (the tool's default) and one signed-K (naive: WGCNA::cor without
`abs()`).

```{bash phase1_absK}
./sim/run_r.sh build_cor_matrix.R \
  --expression_ref /data/Reference_Expression_Whole_Blood.txt.gz \
  --pos /data/Whole_Blood.pos \
  --n_cores 10 \
  --output sim/output/panel_cache/Whole_Blood
# -> sim/output/panel_cache/Whole_Blood.CorMat.RDS  (~28 s)
```

```{bash phase1_signedK}
./sim/run_r.sh sim/build_signed_cor.R \
  --expression_ref /data/Reference_Expression_Whole_Blood.txt.gz \
  --pos /data/Whole_Blood.pos \
  --n_cores 10 \
  --output sim/output/panel_cache/Whole_Blood
# -> sim/output/panel_cache/Whole_Blood.CorMatSigned.RDS  (~40 s)
# Cross-check: identical rownames, dims, and block partition to abs-K.
# NB: sparsity patterns can differ because per-block nearPD is not sign-invariant.
```

**Non-obvious property found here:** the stored `dgCMatrix` is asymmetric — only one
triangle carries some off-diagonal entries. The tool's consumers (`chol`,
`eigen(symmetric=TRUE)`) silently symmetrise on read, so it's harmless in the tool itself
but any code that reads K entrywise must symmetrise explicitly.

# Phase 2: generator + validation

## Cache the per-block Cholesky and eigendecomposition once per panel

```{bash phase2_prep}
./sim/run_r.sh sim/build_panel_prep.R \
  --cor_signed sim/output/panel_cache/Whole_Blood.CorMatSigned.RDS \
  --output sim/output/panel_cache/Whole_Blood.prep.RDS   # ~11 s
```

`panel_prep()` stores `list(gene_universe, block_index, chol_list, eigen_list, ...)` and
symmetrises each K-block via `(K + t(K))/2` before Cholesky/eigen. This is safe for the
MVN sampler because the sampler cares about the closest symmetric approximation.

## Draw one dataset from the harness (optional, standalone sanity check)

```{bash phase2_sim}
./sim/run_r.sh sim/simulate_twas.R \
  --panel_prep sim/output/panel_cache/Whole_Blood.prep.RDS \
  --template_twas /data/twas_results \
  --injection null --seed 42 \
  --output /tmp/sim_smoke_null.tsv
```

## Run the validation gate

```{bash phase2_validate}
./sim/run_r.sh sim/validate_generator.R
# -> sim/output/validate/generator.pdf, generator.log
```

Reads the gate log and asserts three checks:

- **Null marginal**: pooled TWAS.Z is $\mathcal{N}(0,1)$ (KS p > 0.001).
- **Block correlation recovery**: empirical `cor(Z_null)` per block tracks $K_\text{signed}$
  to within theoretical sampling error. Uses a z-scored residual with the correlation-SE
  formula $(1-\rho^2)/\sqrt{n-1}$ — a scale-invariant check, so max-|diff| doesn't blow up
  on large blocks.
- **Injection sanity**: strong positive $\beta$ recovers a positive `cor(z, t)`.

```{r phase2_log_summary, eval=TRUE}
lg <- readLines("output/validate/generator.log")
cat(paste(tail(lg, 12), collapse = "\n"))
```

# Phase 3: in-session core + parity + smoke

## The in-session core

`sim/R/in_session_core.R::score_dataset()` ports `TWAS-GSEA-fast.R` lines 174-419 into a
pure R function so the runner can score thousands of simulated datasets from one R process,
skipping R startup, `.CorMat.RDS` load, and per-block eigen recomputation.

Two subtleties that had to be replicated bit-for-bit for parity:

1. The tool feeds **asymmetric K** to both `eigen(symmetric=TRUE)` (reads lower) and to
   `Cholesky(forceSymmetric(V_hat))` (reads upper). `score_dataset()` therefore does NOT
   symmetrise K — it uses the raw dgCMatrix from `.CorMat*.RDS`.
2. The tool drops duplicate `Alt_ID` rows, which quietly removes rows with NA Alt_ID
   (e.g. genes with no Entrez in the ENSG→Entrez map, ~985 of 7,813 on Whole_Blood).
   The runner replicates this dedup step upstream.

## Parity gate

```{bash phase3_parity}
./sim/run_r.sh sim/parity_check.R
```

7 scenarios spanning {null, signal} × {property, gmt} × {abs-K, signed-K} ×
{one-sided, two-sided} × {probit, directional}. Each simulated dataset is scored by both
paths and P/Estimate/N_Mem_Avail are compared.

```{r phase3_parity_table, eval=TRUE}
p <- fread("output/parity/parity_summary.tsv")
knitr::kable(p, digits = c(NA, NA, 3, 3, 0, 0, 0, 1, 1))
```

Result: worst `|dP|` = 2.63e-7 (one seed near a Brent boundary), typical 1e-13 to 1e-15;
worst `|dEst|` = 5.9e-9. Gate at `|dP| ≤ 1e-5` because the CLI's default REML tolerance is
1e-6 so P is only defined to that precision.

## Smoke: timing + projection

```{bash phase3_smoke}
./sim/run_r.sh sim/smoke.R --n_null 20 --n_signal 20 --n_cores 10
# 6 arms x 20 reps at 10-way = 124 s wall.
# Per-rep wall times: cmap 3.5 s, gmt 9.8 s, ternary 2.4 s, sig_* ~1 s.
```

# Phase 4: full grid, aggregation, plots

## Scenario grid

Written by `sim/write_scenarios.R`; 99 scenarios in three groups:

- **T1E on real drug panels** (property mode): CMAP (3083 drugs) and ternary (2274 drugs)
  under {magnitude/probit + abs-K, directional-2s + abs-K, directional-2s + signed-K},
  1000 reps each.
- **T1E on the full c2 gmt** (set mode): {magnitude, directional-1s + abs-K,
  directional-2s + signed-K}, 500 reps each.
- **Directional headline**: fixed synthetic signed signature $t$ (sparsity 0.05,
  signature_seed=111). Observations are three transforms of the same $t$:
  - `sig_cont` = $t$ + Gaussian noise (SD 1)
  - `sig_tern` = $\text{sign}(t) \cdot \mathbb{1}[|t| > \tau]$
  - `sig_bin`  = $\mathbb{1}[|t| > \tau]$

  Sweep 3 forms × 3 modes (directional 2-sided, directional 1-sided, magnitude) ×
  2 K matrices × 5 $\beta$ levels $\{-0.4, -0.2, 0, 0.2, 0.4\}$ = 90 cells × 300 reps.

```{bash phase4_scenarios}
./sim/run_r.sh sim/write_scenarios.R
# -> sim/scenarios.tsv (99 rows)
```

## Run the grid

```{bash phase4_grid}
./sim/run_r.sh sim/run_grid.R \
  --scenarios sim/scenarios.tsv \
  --outdir sim/output/grid \
  --n_cores 10
# 34,500 tasks at 10-way parallelism -> 2h56m wall (projection was 99 min;
# overshoot from mc.preschedule=FALSE fork overhead at scale).
# -> sim/output/grid/<scenario_id>.long.rds x 99 files
```

## Aggregate and plot

```{bash phase4_agg}
./sim/run_r.sh sim/aggregate.R
./sim/run_r.sh sim/plots.R
# -> sim/output/aggregate/scenario_summary.tsv   (99 rows)
# -> sim/output/aggregate/power_summary.tsv      (72 rows)
# -> sim/output/plots/fig1_t1e_qq.pdf .. fig5_absK_vs_signedK.pdf
```

# Results

## T1E on real property panels + gmt

```{r res_t1e_real, eval=TRUE}
s <- fread("output/aggregate/scenario_summary.tsv")
t1e_real <- s[grepl("^t1e_", scenario_id), .(scenario_id, arm, feature_source, n_reps,
                                                n_features,
                                                `frac P<0.05` = round(frac_p_lt_05, 4),
                                                `FWER@0.05`   = round(fwer_at_05,   3),
                                                `mean #FD`    = round(mean_n_fd,    2))]
knitr::kable(t1e_real[order(feature_source, arm)])
```

**The abs-K arms (directional AND magnitude/probit) are correctly calibrated** at the
nominal 5% per-test rate and ~5% family-wise error rate.

**The directional + signed-K combination catastrophically inflates T1E** — per-test rate
0.19-0.27 (4-5× nominal), FWER 0.85-1.00, with hundreds of false discoveries per
replicate (768 out of 7129 gmt sets in the worst case). This is a systematic misspecification,
not sampling noise (n_reps = 500 to 1000 per scenario).

## T1E on the directional-headline null cells ($\beta=0$)

```{r res_t1e_dh, eval=TRUE}
dh0 <- s[grepl("^dh_", scenario_id) & beta == 0,
          .(feature_source, arm, cor_matrix,
            `frac P<0.05` = round(frac_p_lt_05, 4))]
knitr::kable(dh0[order(feature_source, arm, cor_matrix)])
```

Same pattern replays across all three feature forms (sig_cont, sig_tern, sig_bin):
abs-K arms are calibrated; signed-K + directional is 0.22-0.44 vs nominal 0.05.
**Crucially, signed-K + magnitude/probit is fine (~0.05)** — the inflation is a specific
interaction between the signed relmat and the directional response variable.

## Directional headline power

```{r res_power, eval=TRUE}
p <- fread("output/aggregate/power_summary.tsv")
w <- p[beta %in% c(-0.4, 0.4),
        .(feature_source, arm, cor_matrix, beta,
          `mean Est` = round(mean_est, 3),
          `power(P<.05)` = round(power_p05, 3),
          `sign recovery` = round(sign_recovery, 3))]
knitr::kable(w[order(feature_source, arm, cor_matrix, beta)])
```

- **Directional two-sided on abs-K** (the tool's proper setting): power = 1.0 at $|\beta|=0.4$
  for both signs on continuous and signed-ternary properties. Estimates track $\beta$ linearly
  ($\pm 0.28$ at $\pm 0.4$). Sign recovery = 1.0.
- **Directional one-sided on abs-K**: power = 1.0 at $\beta=+0.4$, 0.0 at $\beta=-0.4$ —
  reversal missed by design.
- **Magnitude on signed properties**: near-zero power. Correct — $|Z|$ is direction-blind
  and the injected mean has expected magnitude ~0.
- **Magnitude on unsigned binary property**: power = 1.0 — unsigned enrichment picked up
  correctly because the annotated genes have larger $|Z|$ regardless of sign.
- **Directional on unsigned binary property**: sign_recovery = 0 (reversal). This is an
  artefact of the specific `signature_seed=111` — the 391 rnorm draws happen to have
  non-zero empirical mean, aligning Estimate against $\beta$. A single-seed study exposes
  the sensitivity but can't make a general claim; averaging across seeds would be needed.

## Signed-K vs abs-K power at matched $|\beta|$

```{r res_absK_vs_signedK, eval=TRUE}
w2 <- dcast(p[beta == 0.4 & !is.na(power_p05)],
             feature_source + arm ~ cor_matrix, value.var = "power_p05")
setnames(w2, c("feature_source","arm","power_absK","power_signedK"))
w2[, delta := round(power_signedK - power_absK, 3)]
knitr::kable(w2)
```

Signed-K gives higher raw power on the unsigned-binary case, but that "power" is
uninterpretable given the T1E inflation shown above. On the properly-calibrated
directional arms (sig_cont, sig_tern), abs-K and signed-K reach the same 1.0 power at
$|\beta|=0.4$, so signed-K offers no real gain — only a calibration cost.

## Figures

The plots directory holds:

- `fig1_t1e_qq.pdf` — QQ of pooled P per T1E arm (nine panels).
- `fig2_dh_power_vs_beta.pdf` — power curves vs $\beta$ by (feature × arm × K).
- `fig3_dh_sign_recovery.pdf` — fraction of significant reps with correct-sign Estimate.
- `fig4_discretisation_penalty.pdf` — power(sig_cont) - power(sig_tern) at matched $\beta$.
- `fig5_absK_vs_signedK.pdf` — power($K$=signed) - power($K$=abs) per arm.

# Discussion

## Practical implication

The shipped tool's default `abs(K)` relmat is correctly calibrated under both directional
and magnitude modes. A naive drop-in switch to signed $K$ (i.e. removing the `abs()` in
`R/build_cor_matrix_helper.R:40`) is **not safe**: T1E inflates 4-5× and FWER approaches 1.
This is a specific interaction between the signed relmat and the directional response
variable — signed $K$ under magnitude/probit remains calibrated.

## Limitations

- **Single tissue / single panel.** Everything is Whole_Blood, 7,813 genes, 503 samples.
  The magnitude of inflation could change with different block-size distributions or
  different sample sizes.
- **Single signature realisation** for the directional headline (signature_seed = 111).
  A properly general claim about e.g. "unsigned-binary → null directional" needs multiple
  seeds.
- **Synthetic signatures only** in the headline arm. The "realism" arm proposed in the
  original spec (inject via a real drug signature, recover via continuous vs ternary
  discretisation) was not run.
- **Fixed sparsity / $\sigma_{obs}$ / $\tau$** for the sig_* forms — no sweep over these.
- **CMAP / ternary property panels** were used only for the T1E-real block; the
  power side wasn't run against real drug signatures either.
- The `dgCMatrix` produced by `build_cor_matrix.R` is stored asymmetrically; while
  `score_dataset()` reproduces the tool's exact numerical behaviour on this asymmetry,
  a downstream user would be safer if the tool wrote symmetric K.

## Future directions

- **Root-cause the signed-K T1E inflation.** The most likely culprit is the interaction
  between negative $K$ eigenvalues (after nearPD-repair) and the profiled REML
  likelihood's boundary behaviour. A per-block simulation with known signed-K covariance
  would isolate whether the issue is REML or the GLS whitening.
- **Average across signature seeds** in the directional headline to give the sig_bin
  arm a fair null test.
- **Add a realism arm** using real drug signatures as $t$, then testing continuous vs
  ternary discretisation of the same signature.
- **Sweep sparsity and $\sigma_{obs}$** to characterise discretisation penalty as a
  function of the ratio of true-signal to noise.
- **Multi-panel comparison** (e.g. brain-tissue panels with different block-size
  distributions) to test whether the abs-K calibration is panel-general.
- **Fix `build_cor_matrix_helper.R`** to write a symmetric `dgCMatrix` regardless
  of the sparse_struc indexing quirk. Silently harmless today, potentially confusing
  tomorrow.
